\Large{All in the family:\\German twin finishing times in\\the 2016 women's Olympic marathon}

### 2017-02-03 paper/paper.tex
ready for david final editing

At 9:30am on August 14, 2016, the Women's Olympic marathon kicked off
in Rio de Janeiro, Brazil, when 156 runners from 80 countries across
the world left the starting line en route to their destination 42.195
kilometers away. Two hours, twenty-four minutes, and four seconds
later, Jemima Sumgong of Kenya would be the first to cross the
finishline and take home gold; Sumgong was just three and one-half
minutes slower than her prior personal best time in the
marathon. Approximately 21 minutes later, twin marathoners from
Germany, Anna and Lisa Hahner, would cross the finishline together,
holding hands and celebrating a personal victory. Although the Hahners
would finish 81st and 82nd, respectively, well behind the winners of
the marathon, Anna Hahner would describe their joint finish as a
``magical moment.''

@@

The media quickly picked up on the Hanher story as an image of the
beaming twins finishing hand-in-hand captured a public audience. While
many believed the moment was a reflection of the Olympic spirit, not
everyone agreed with this rosy interpretation. The twins' happy facial
expressions at the finish were portrayed as a bit contrived---smiling
like ``Honigkuchenpferde,'' cookies in the shape of a horse---and the
sports director of the German Athletics Federation, Thomas
Kurschilgen, stirred up controversy when he suggested that the
Hahners' photo-finish was no coincidence.  Kurschilgen averred that
the twins slowed down so as to finish simultaneously and create a
spectacle which would ``generate media attention.''  Kurschilgen
justified his charge with the fact that the twins ran in the Rio
marathon at least 18 minutes slower their personal best times prior to
the
Olympics.\footnote{\url{https://www.nytimes.com/2016/08/17/sports/olympics/twins-finish-marathon-hand-in-hand-but-their-country-says-they-crossed-a-line.html}}
Not surprisingly, Kurschilgen's accusations were denied by the Hahner
twins, who claimed that their simultaneous finish was simply an
unintended coincidence.

@@

What happened in the women's Olympic marathon, and how might we
develop a statistical approach that assesses whether the Hahner twin's
finish was coincidental or intentional?  These two interpretations are
clearly at odds. If the former, then the Hahners are to be celebrated
and their finish treated as an expression of the spirit behind the
Olympic games. If the latter, though, then the twins may have violated
this spirt by not trying hard enough. It is perhaps too easy for us to
write such a glib sentence---neither of us can fathom being able to
complete a marathon anywhere in the vicinity of two and a half
hours---but we nonetheless want to know what the data from the Olympic
marathon tell us.  Was the Hahner finish in the 2016 women's
Olympic marathon a lovely coincidence or something else?

Among female Olympic marathoners, the Hahner twins were not alone in
their familial ties.  The marathon also featured twins from North
Korea, Kim Hye-song and Kim Hye-gyong, who posted identical times and
finished 10th and 11th in the race, respectively. The Kim finish,
unlike the Hahner finish, appears devoid of post-race
controversy. Moreover, three triplets from Estonia competed in the Rio
marathon, although only two, Lily Luik and Leila Luik, finished it, in
97th and 114th place, respectively. The third Estonia triplet, Liina
Luik, recorded what is known as a DNF---an abbreviation that means did
not finish.  Although our focus here is the Hahner twins, we touch on
the Kim twins and Luik triplets in the course of what follows.

@@

The Kurschilgen accusation against the Hahner twins has two
components, that these two women ran slowly \emph{and} that they
finished simultaneously.  We suspect that Kurschilgen would not have
expressed ire at the Hahners had they finished in 1st and 2nd place in
Rio, hand-in-hand with wide grins, but of course we do not know this.
As such, our investigation of the charges that Kurschilgen offered
will distinguish between a slow finish versus a simultaneous finish.

Our research design is twofold.  First, we present visualizations that
describe various features of the Hahner twin's results, and an
important element in our visualizations is the difference between a
runner's Rio time and her prior personal best time in the marathon.
Overall, our visualizations suggest that the Hahner twins' pace in the
marathon was slow, albeit not excessively so, but that their
simultaneous finish was quite unusual given the twins' differences in
abilities (and similarly for the Kim twins).  We then turn to a
regression-based simulation of the marathon, which we develop based on
relationships across the Olympic marathoners between personal best
times and observed Rio finishing times.  Our simulations show that the
Hahner twins did not run appreciably slowly yet finished suspiciously
close to each other given the disparty between their personal best
times.  We return in our conclusion to Kurschilgen's claims about the
Hahners and offer our thoughts about their validity.

@@

Figure \ref{fig:scatter} and Figure \ref{fig:scatterhalf} contain two plots, both of which describe
how finishing times from the Rio Olympic women's marathon varied as a
function of athletes' personal best times and and as function of their Rio half marathon
split times.  The points in the plots are colored by twin/triplet
status, and both plots contain second-order polynomial regression
lines.  We treat an athlete's prior personal best marathon time as a
measure of the athlete's underlying marathon talent (seen in Figure \ref{fig:scatter}), and we interpret
an athlete's half marathon split as a measure of the athlete's skill
at marathon running on August 14, 2016, the date of the women's
marathon in Rio (seen in Figure \ref{fig:scatterhalf}).

\begin{figure}[!ht]
  \caption{Olympic results as a function of ability measured by personal best}
  \label{fig:scatter}
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot.pdf}
    \caption{} 
    \label{fig:45degreeplot}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{studentized_residuals.pdf}
    \caption{}
    \label{fig:studentizedresiduals}
  \end{subfigure}
\end{figure}

\begin{figure}[!ht]
  \caption{Olympic results as a function of ability measured by half marathon split}
  \label{fig:scatterhalf}
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot_half.pdf}
    \caption{}
    \label{fig:45degreeplot_half}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{studentized_residuals_half.pdf}
    \caption{}
    \label{fig:studentizedresidualshalf}
  \end{subfigure}
\end{figure}

Figure \ref{fig:45degreeplot} includes a solid 45-degree line as
the axes in the figure correspond to full marathons; in addition, the
figure's rug marks denote personal best times of runners who DNFed.
Given the paucity of points (five of them) below the pictured
45-degree line, the vast majority of Olympic marathoners ran slower in
Rio compared to their personal bests.  Relative to personal best
times, the Hahner twins were definitely on the slow side, but a number
of runners had greater differences between their Olympic times and
their personal bests than the Hahners.  These runners are denoted with
squares in Figure \ref{fig:45degreeplot}, and there are 13 such
symbols, highlighting approximately 10\% of the racers who completed
the marathon.  The figure shows moreover that there were two women who
had personal best times slightly faster than the Hahner's and yet
finished after the German twins.  Although Figure
\ref{fig:45degreeplot} suggests that the Hahner twins were slower than
one would have expected given their previous best marathon times, it
is not consistent with the accusation that they dramatically slowed
down in the Rio marathon.

@@

One limitation of using personal best times as indicators of
underlying athletic talent is that these times are potentially
confounded by the marathon courses at which they were set (some
courses, like the Berlin marathon, are known for fast times) and race
conditions like weather.  In addition, personal best times may not
capture raceday idiosyncrasies that might affect individual runners.
With that in mind, Figure \ref{fig:45degreeplot_half} plots marathon
finishing times against half marathon splits.  This figure has a
regression line as before but no 45-degree line.

Figure \ref{fig:45degreeplot_half} shows that there was nothing
abnormal about the Hahner twins' overall finishing times, conditional
on their half marathon split times.  This is consistent with the
previous Figure \ref{fig:45degreeplot} and inconsistent with
accusations made against the Hahner twins, at least the part of the
accusation that focused on their  overall pace.

If the Hahner twins did not slow down excessively, might they have
run somewhat strategically at the end of the race in order to generate
a simultaneous finish?  We now offer a visualization that speaks to
this question.

The personal best times of the Hahner twins were 115 seconds apart and
their official finishing times were separated by one second.  Is such
a 115 to one compression typical among pairs of runners or, in
contrast, was it unusual?  Are there other pairs of marathoners who
had a difference between personal best times of 115 seconds apart and,
if so, how close where their finishing times?

@@

\begin{figure}[!ht]
 \caption{Differences in personal best times and finishing times}
  \label{fig:diffdiff}
 \centering
 \begin{subfigure}{.49\textwidth}
 \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot.pdf}
 \caption{}
  \label{fig:diffdiffscatter}
 \end{subfigure}
  \begin{subfigure}{.49\textwidth}
 \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot_half.pdf}
 \caption{}
  \label{fig:diffdiffscatterhalf}
 \end{subfigure}
\end{figure}

@@

Consider first the Hahner twins.  They are effectively located on the
horizontal axis because their difference in finishing times is one
second.  However, there are many points about the Hahner's orange dot,
and this shows that, conditional on an approximate 115 second
difference in personal best times, most marathoners did not have close
finishing times like the Hahners.  Some pairs of runners with around
115 second personal best differences had finishing time differences of
1000 seconds, i.e., in excess of 15 minutes.  Of course the points in
Figure \ref{fig:diffdiffscatter} are not independent, but they provide
a sense of the dispersion in finishing time differences that one can
expect conditional on differences in personal best times.

@@

To get a sense of the lockstep nature of the Kim twins race, Figure
\ref{fig:secondsbehind} describes each runner's status at the various
split times on the marathon course.  Each dot in the figure depicts a
recorded split and the number of seconds each runner was behind the
race leader at the time.  There are more dots at earlier splits due to
the accumulation of DNFs.  The usual color scheme applies here, and
the Estonian DNF prior to the half marathon split is evident.  The
orange dots in the figure represent the Hahner twins, and as noted in
the figure Lisa Hahner was ahead of her sister through 15 kilometers
at which point Anna surged.  Figure \ref{fig:secondsbehind} contains
two red dots representing the North Korean Kim twins, but this is not
visually apparent because the Kim twins had identical split times
during the entire race marathon.

\begin{figure}[!ht]
  \centering
  \caption{Runner status by split}
  \label{fig:secondsbehind}
  \includegraphics[scale = 1]{seconds-behind.pdf}
  \begin{flushleft}
    \emph{Note: each dot represents one runner at a split.  DNFs are
      not pictured, and splits are not to scale.}
  \end{flushleft}
\end{figure}

@@

We now consider the following question: if we take into consideration
the Hahner twin's similarities as runners as well as natural variation
in marathon finishing times, what is the probability that the twins
would finish the Rio race at roughly the same time and/or
sequentially?  To answer this question we need to know the
counterfactual distribution of potential marathon finishing times that
would have occurred had the Hahner twins independently (in particular,
of each other) and repeatedly run the Rio marathon, holding constant
marathon conditions, the abilities of other runners, and so forth.
Access to such a distribution would establish the set of potential
outcomes that could have occurred on August 14, 2016, and we could in
principle use this distribution to determine the likelihood that a
simultaneous finish by the Hahners, or at least a near-simultaneous
finish, occurred by chance alone.  If these twins rarely finish the
marathon together in such a counterfactual world, then one might be
skeptical that their observed finish occurred without
coordination---as they in fact claimed.

Unfortunately for us but nonetheless fortunately for the race's
participants, it is not possible to rerun the women's Olympic marathon
in Rio to establish a distribution of potential race outcomes for the
Hahner twins.  However, we can attempt to simulate this distribution
by estimating the distribution of every other runner's finishing time,
conditional on prior abilities, and drawing from that distribution in
order to observe the likelihood that, for example, Lisa and Anna
Hahner finished simultaneously.

To estimate the conditional distribution of each Rio final results, we
assume that each runner's marathon time \Yi\ is distributed normally
with a mean that is a quadratic function of her personal best time
$X_i$:

$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i} + \beta_2\,X^2_{i}  + e_i, \sigma\right).$$
%%
where $N\left(\cdot,\cdot\right)$ denotes a normal distribution.
Personal best times are all prior to the Rio marathon.  We estimate
$\beta_1$ , $\beta_2$, and $\sigma$ using ordinary least squares and
all finishing runners except for the twins and triplets.  For a
simulated marathon, we then draw a runner's time from an estimated
distribution, conditioned on the runner's personal best marathon time.
Once a race is simulated for all runners, twins and triplets included,
we locate Anna and Lisa Hahner in our set of simulated results and
record both the time between their simulated finishes and the
difference in their simulated ranks.  We then simulate a new
race---drawing a new set of finishing times---and record the same
quantities.  In particular, the steps of the simulation are as
follows.

\begin{enumerate}
\item Ignoring twins and triplets, estimate with least squares a
  linear model that predicts a runner's finishing time $Y_i$ based on
  her pre-Olympic personal best time $X_i$ and her best time squared.
\item Extract the resulting coefficient vector, estimated covariance
  matrix, and estimated regression variance from this model.
\item For each simulated rate, draw intercept and slope estimates
  $\tilde{\beta}_0$, $\tilde{\beta}_1$, and $\tilde{\beta}_2$,
  respectively, from a multivariate normal distribution with mean
  equal to the previously estimated coefficient vector and covariance
  equal to the previously estimated covariance matrix.
\item Given values of $\tilde{\beta}_0$, $\tilde{\beta}_1$, and
  $\tilde{\beta}_2$, for each runner draw an error $\tilde{e}$ from a
  normal distribution with mean zero and a standard deviation equal to
  the standard deviation of the original regression model's residuals.
\item For each runner, predict her final result by combining the
  randomly generated beta coefficients and error terms,
  $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +
  \tilde{\beta}_2\,X_i^2 + \tilde{e}$.
\item Eliminate each runner from the simulated race with a probability
  equal to the observed fraction of marathoners who did not finish the
  marathon.
\item Calculate the difference in time and difference in ranking
  between Anna Hahner and Lisa Hahner, assuming both finished the
  race.
\item Repeat above steps 10,000 times.
\end{enumerate}

David---comments here?    Or above?  This is not from the simulation
--- so perhaps move to top of section??

From our simulations we can generate intervals which describe the
extent of the variability in racer finishing times.  In 95\% of the
simulations in which Anna Hahner completed the marathon, her finishing
time was between 9,480 seconds and 10,703 seconds.  This is consistent
with Anna's observed finishing time, which was 9,932 seconds.  In
fact, our simulations estimate that Anna should have finished the Rio
marathon in 10,092 seconds, which is slightly more than two and
one-half minutes slower than her actual time.  Lisa Hahner's
corresponding interval is 9,427 to 10,619 with an observed finishing
time of 9,933 seconds.

The bottom line here is that our simulated marathon finishes are
consistent with the Hahner twins' finishing times.  And note that the
regression model underlying the simulations was estimated without the
Hahner twins (and same for the Kim twins and Luik triplets).
According to the our simulations, then, neither Hahner twin ran
appreciably slowly, conditional on personal best times prior to the
Rio Olympics.

@@

The results plotted in the histograms raise questions about the
credibility of Anna and Lisa Hahner's story.  In particular, the
histogram suggest that a simultaneous finish would be very rare if
Anna and Lisa had run independently of each other.  For example, in
fewer than 300 of the 10,000 simulated races did Anna and Lisa
finished within 30 seconds of each other.  Moreover, they finished in
consecutive rank in fewer than 175 of the 10,000 races.  Therefore,
the close finish that was observed, where Anna and Lisa crossed the
finish line one-after-the-other, would be highly unlikely if the two
German twins had raced independently of each other.

Earlier we noted that personal best times might arguably be poor
measures of runner abilities.  With this in mind, we repeated our
simulations but replaced personal best times with half marathon
splits.  See Figure \ref{fig:simdiffhalf}.  While this change reduces
the variation in the predicted outcome of each runner and therefore,
reduces the expected distance between Anna and Lisa Hahner, even with
half marathon split as a measure of ability it is still quite rare for
Anna and Lisa to finish simultaneously or consecutively.

@@

An important point regarding our simulations is the way that we
handled DNFs.  As our description above indicates, we assumed that DNF
probabilities are the same for all runners and that the likelihood of
a DNF is a not function of a runner's anticipated marathon finishing
time.  The rug marks in Figure \ref{fig:45degreeplot} suggest that
runners with better personal best times may be more likely to DNF than
other runners, \emph{ceteris paribus}.  We suspect that this occurs
because some better runners may expend excessive energy trying to
achieve a good result in the marathon and in so doing injure or
exhaust themselves; lesser runners, in contrast, may be content to
finish respectably.  Regardless of the validity of this conjecture,
Figure \ref{fig:45degreeplot} shows that the Hahner twins are
representative of sort of runners who DNFed in Rio.  Since our
simulated Hahner statistics are conditioned on both Hahner twins
finishing, it follows that they are conservative.  The fact that both
women finished the marathon was notable in and of itself, and by
discounting the possibility of a Hahner DNF we are giving the benefit
of the doubt to the Hahner twins.

>>>>

At 9:30am on August 14, 2016, the women's Olympic marathon kicked off
in Rio de Janeiro, Brazil, when 156 runners from 80 countries across
the world left the starting line en route to their destination 42.195
kilometers away. Two hours, twenty-four minutes, and four seconds
later, Jemima Sumgong of Kenya would be the first to cross the
finishline and take home gold; Sumgong was just three and one-half
minutes slower than her prior personal best time in the
marathon. Approximately 21 minutes later, twin marathoners from
Germany, Anna and Lisa Hahner, would cross the finishline together,
holding hands and celebrating a personal victory. Although the Hahners
would finish 81st and 82nd, respectively, well behind the winners of
the marathon, Anna Hahner would describe their joint finish as a
``magical moment.''

@@

The media quickly picked up the Hanher story as an image of the
beaming twins finishing hand-in-hand captured a public audience. While
many believed the moment was a reflection of the Olympic spirit, not
everyone agreed with this rosy interpretation. The twins' happy facial
expressions at the finish were portrayed as a bit contrived---smiling
like ``Honigkuchenpferde,'' cookies in the shape of a horse, was the
description offered by one editorialist---and the sports director of
the German Athletics Federation, Thomas Kurschilgen, stirred up
controversy when he suggested that the Hahners' photo-finish was no
coincidence.  Kurschilgen averred that the twins slowed down so as to
finish simultaneously and create a spectacle which would ``generate
media attention.''  Kurschilgen justified his charge with the fact
that the twins ran the Rio marathon over 18 minutes slower their
personal best times prior to the Olympics.  Not surprisingly,
Kurschilgen's accusations were denied by the Hahner twins, who claimed
that their simultaneous finish was simply an unintended coincidence.

@@

What happened in the women's Olympic marathon, and how might we
develop a statistical approach that assesses whether the Hahner twin's
finish was coincidental or intentional?  These two interpretations are
clearly at odds. If the former, then the Hahners are to be celebrated
and their finish treated as an expression of the spirit behind the
Olympic games. If the latter, though, then the twins may have violated
this spirt by not trying hard enough. It is perhaps too easy for us to
write such a glib sentence---neither of us can fathom being able to
complete a marathon anywhere in the vicinity of two and a half
hours---but we nonetheless want to know what the data from the Olympic
marathon tell us.

%% Was the Hahner finish in the 2016 women's Olympic marathon a lovely coincidence or something else?

Among female Olympic marathoners, the Hahner twins were not alone in
sharing familial ties.  The Rio marathon also featured twins from
North Korea, Kim Hye-song and Kim Hye-gyong, who posted identical
times and finished 10th and 11th in the race, respectively. The Kim
finish, unlike the Hahner finish, appears devoid of post-race
controversy. Moreover, three triplets from Estonia competed in the Rio
marathon, although only two, Lily Luik and Leila Luik, finished it, in
97th and 114th place, respectively. The third Estonia triplet, Liina
Luik, recorded what is known as a DNF---an abbreviation that means did
not finish.  Although our focus here is the Hahner twins, we touch on
the Kim twins and Luik triplets in the course of what follows.

@@

The Kurschilgen accusation against the Hahner twins has two
components, that these two women ran slowly \emph{and} that they
finished simultaneously.  We suspect that Kurschilgen would not have
expressed ire at the Hahners had they finished in 1st and 2nd place in
Rio, hand-in-hand with wide grins, but of course we cannot be certain
of this.  Thus, our investigation of the charges that Kurschilgen
offered will distinguish between a slow finish versus a simultaneous
finish.

Our research design is twofold.  First, we present visualizations that
describe various features of the Hahner twin's results, and an
important element in our visualizations is the difference between
runners' Rio times and their prior personal best marathon times.  Our
visualizations suggest that the Hahner twin's pace in the marathon was
slow, albeit not excessively so, but that their simultaneous finish
was quite unusual given the twins' differences in abilities (and
similarly for the Kim twins).  We then turn to a regression-based
simulation of the marathon, which we develop based on relationships
across the Olympic marathoners between personal best times and
observed Rio finishing times.  Our simulations reinforce what we
observed in visualization, namely, that the Hahner twins did not run
appreciably slowly yet finished suspiciously close to each other.  We
return in our conclusion to Kurschilgen's claims about the Hahners and
offer thoughts about their validity.

@@

One way that we might assess whether an athlete's Rio finish was
unusual---or, say, whether two athletes' finishes were jointly
unusual---is by comparing observed finishing times on August 14 to a
measure of underlying marathon talent.  For the latter we use prior
personal best marathon times.  This is a natural measure of marathon
talent, we believe, but it is not entirely free of complications.  One
limitation of using personal best times as indicators of underlying
marathon skill is that these times are potentially confounded by the
marathon courses at which they were set; some courses, like the Berlin
marathon, are known for relatively fast times.  In addition, personal
best times may be confounded by conditions, like weather, that varied
across races.  Finally, personal best times may not capture Olympic
raceday idiosyncrasies that could affect individual runners.  For
example, a runner could have woken up with a minor cold on the morning
of August 14, 2016.  With these concerns in mind, we use half marathon
split times from the Olympic marathon as a secondary measure of
athletic skill in marathoning.  Our two measures of marathon talent
are featured in the visualizations below.

Figure \ref{fig:scatter} contains two plots, both of which describe
how finishing times from the Rio women's marathon varied as a function
of athletes' personal best times and as function of their half
marathon split times.  The points in the plots are colored by
twin/triplet status, and both plots contain second-order polynomial
regression lines.  Rug marks in both plot indicate times (either
personal best or half marathon split) of runners who earned a DNF
result.

\begin{figure}[!ht]
  \centering
  \caption{Olympic finishing times as a function of underlying marathon talent}
  \label{fig:scatter}
  \begin{subfigure}{.495\textwidth}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot.pdf}
    \caption{Based on personal best} 
    \label{fig:45degreeplot}
  \end{subfigure}
  \begin{subfigure}{.495\textwidth}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot_half.pdf}
    \caption{Based on half marathon split} 
    \label{fig:45degreeplot_half}
  \end{subfigure}
\end{figure}

Considering first the relationship between Olympic finishing and
personal best times, Figure \ref{fig:45degreeplot}'s solid 45-degree
line is informative.  Given the paucity of points (only five of them)
below this line, it follows that the vast majority of Olympic
marathoners ran slower in Rio compared to prior personal bests.
Relative to personal best times, the Hahner twins were definitely on
the slow side, but a number of runners had greater differences between
their Olympic times and their personal bests than the Hahners.  These
runners are denoted with squares in Figure \ref{fig:45degreeplot}, and
there are 13 such symbols, highlighting approximately 10\% of the
racers who completed the Rio marathon.  Moreover, the figure shows
that there were two women who had personal best times slightly faster
than the Hahner's and yet finished after the German women.  Although
Figure \ref{fig:45degreeplot} suggests that the Hahner twins were
slower than one would have expected given their previous best marathon
times, it is not consistent with the accusation that they dramatically
slowed down in the Rio marathon.

@@

With respect to our second measure of marathon skill, Figure
\ref{fig:45degreeplot_half} shows that there was nothing abnormal
about the Hahner twins' overall finishing times, conditional on their
half marathon split times.  This is consistent with the previous
Figure \ref{fig:45degreeplot} and inconsistent with accusations made
against the Hahner twins, at least the part of the accusation that
focused on their overall pace.

Figure \ref{fig:residuals} is a visualization of how unusual, or not,
the Hahner and Kim twins' finishing times were in the Olympic
marathon, conditional on our two measures of marathon talent.  The
left and right panels of the figure are based on the polynomial
regressions in Figure \ref{fig:scatter}'s panels, respectively.  We
calculate residuals from Figures \ref{fig:45degreeplot} and
\ref{fig:45degreeplot_half}, we Studentize the residuals, and then we
plot the ordered residuals in Figure \ref{fig:residuals}.

\begin{figure}[!ht]
  \centering
  \caption{Studentized residuals from Olympic finishing time regressions}
  \label{fig:residuals}
  \begin{subfigure}{.495\textwidth}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{studentized_residuals.pdf}
    \caption{Based on personal best} 
    \label{fig:studentizedresiduals}
  \end{subfigure}
  \begin{subfigure}{.495\textwidth}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{studentized_residuals_half.pdf}
    \caption{Based on half marathon split}
    \label{fig:studentizedresiduals_half}
  \end{subfigure}
\end{figure}

The cumulative residuals depicted in Figure \ref{fig:residuals} are
consistent with our interpretation of the Hahner twin's Olympic
finishing times as not particularly remarkable.  With personal best
time as a measure of marathon talent, the two Hahner residuals are in
the right tail of the residual distribution; however, their locations
are not extreme.  One residual is located at the 88th percentile and
the other at the 93rd.  Kim twin residuals are similarly not
remarkable although they are in the left tail of the residual
distribution, indicating that the Kims ran faster than one would have
expected.  If one uses half marathon split times as a measure of
marathon talent, similar results obtain.  Neither Hahner twin had a
finishing time that was particularly unusual given her half marathon
split, and this applies to the Kim twins as well.

If the Hahner twins did not slow down excessively, might they have run
somewhat strategically at the end of the Olympic marathon in order to
generate a simultaneous finish?  We now offer a visualization that
speaks to this question.

The personal best times of the Hahner twins were 115 seconds apart and
their official finishing times were separated by one second.  Is such
a 115 to one compression typical among pairs of runners?  Moreover,
are there other pairs of Olympic marathoners who had a difference
between personal best times of 115 seconds apart and, if so, how close
where their finishing times?

@@

\begin{figure}[!ht]
 \caption{Pair-wise differences in Olympic finishing times and
   differences in marathon talent}
  \label{fig:diffdiff}
 \centering
 \begin{subfigure}{.49\textwidth}
 \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot.pdf}
 \caption{Based on differences in personal best times}
  \label{fig:diffdiffscatter}
 \end{subfigure}
  \begin{subfigure}{.49\textwidth}
 \includegraphics[width=\textwidth, keepaspectratio]{diff_in_diff_scatter_plot_half.pdf}
 \caption{Based on differences in half marathon splits}
  \label{fig:diffdiffscatterhalf}
 \end{subfigure}
\end{figure}

@@

Consider first the Hahner twins.  They are effectively located on the
horizontal axis because their difference in finishing times is one
second.  However, there are many points above the Hahner's orange dot,
and this shows that, conditional on an approximate 115 second
difference in personal best times, most marathoners did not have close
finishing times like the Hahners.  Some pairs of runners with around
115 second personal best differences had finishing time differences of
1000 seconds, i.e., in excess of 15 minutes.  Of course the points in
Figure \ref{fig:diffdiffscatter} are not independent, but they provide
a sense of the dispersion in finishing time differences that one can
expect conditional on differences in personal best times.

@@

Similar conclusions follow from Figure \ref{fig:diffdiffscatterhalf},
which plots pair-wise differences between Olympic finishing times and
half marathon split times.  Namely, many pairs of runners had similar
differences in half marathon times as the Hahner and Kim twins, but
the vast majority of these pairs did not have close finishing times.

Figure \ref{fig:secondsbehind} provides a sense of the lockstep nature
of the Kim twins, and it describes each Olympic runner's status at
various split times on the marathon course.  Each dot in the figure
depicts a recorded split and the number of seconds each runner was
behind the race leader at the time.  There are more dots at earlier
splits due to the accumulation of DNFs.  The usual color scheme
applies here, and the Estonian triplet DNF prior to the half marathon
split is evident. As noted in the figure Lisa Hahner was ahead of her
sister through 15 kilometers at which point Anna took the lead.
Figure \ref{fig:secondsbehind} contains two red dots representing the
North Korean Kim twins, but this is not visually apparent because the
Kim twins had identical split times during the entire marathon.

\begin{figure}[!ht]
  \centering
  \caption{Runner status by split}
  \label{fig:secondsbehind}
  \includegraphics[scale = 1]{seconds-behind.pdf}
  % \begin{flushleft}
  %   \emph{Note: each dot represents one runner at a split.  DNFs are
  %     not pictured, and splits are not to scale.}
  % \end{flushleft}
\end{figure}

@@

Our visualizations are useful and they shed a fair bit of light on the
Hahner twin's performance in Rio.  In the interest of increasing
precision, we now pose the following question: if we take into
consideration the Hahner twin's similarities in marathon talent as
well as natural variation in marathon finishing times, what is the
probability that the twins would finish the Rio race at roughly the
same time and/or sequentially?  To answer this question, we need to
know the counterfactual distribution of potential marathon finishing
times that would have occurred had the Hahner twins independently (in
particular, of each other) and repeatedly run the Rio marathon,
holding constant marathon conditions, the abilities of other runners,
and so forth.  Access to such a distribution would establish the set
of potential outcomes that could have occurred on August 14, 2016, and
we could in principle use this distribution to determine the
likelihood that a simultaneous finish by the Hahners, or at least a
near-simultaneous finish, occurred by chance alone.  If these twins
rarely finish the marathon together in such a counterfactual world,
then one might be skeptical that their observed finish occurred
without coordination---as they in fact claimed.

Unfortunately for us but nonetheless fortunately for the race's
participants, it is not possible to rerun the women's Olympic marathon
to establish a distribution of potential race outcomes for the Hahner
twins.  However, we can attempt to simulate this distribution by
estimating the distribution of every other runner's finishing time,
conditional on marathon talent, and then drawing from this
distribution in order to calculate the likelihood that, for example,
Lisa and Anna Hahner finished simultaneously.

To estimate the conditional distribution of each Rio final results, we
assume that each runner's marathon time \Yi\ is distributed normally
with a mean that is a quadratic function of marathon skill $X_i$. Here
we start by assuming that marathon skill can be measured by personal
best marathon time:
$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i} + \beta_2\,X^2_{i}  + e_i, \sigma\right),$$
%%
where $N\left(\cdot,\cdot\right)$ denotes a normal distribution.  We
estimate $\beta_1$ , $\beta_2$, and $\sigma$ using ordinary least
squares and all finishing Rio marathoners except for the Hahner/Kim
twins and Luik triplets.  For a simulated marathon, we draw a runner's
time from an estimated distribution and condition on the runner's
personal best marathon time.  Once a race is simulated for all
runners, twins and triplets included, we record both the time between
Anne and Lisa Hahner's simulated finishes and the difference in their
simulated ranks.  We then simulate a new race---drawing a new set of
finishing times---and record the same quantities. The steps of the
simulation are as follows.
\begin{enumerate}
\item Ignoring twins and triplets, estimate with least squares a
  linear model that predicts a runner's finishing time $Y_i$ based on
  her pre-Olympic personal best time $X_i$ and her best time squared.
\item Extract the resulting coefficient vector, estimated covariance
  matrix, and estimated regression variance from this model.
\item For each simulated rate, draw intercept and slope estimates
  $\tilde{\beta}_0$, $\tilde{\beta}_1$, and $\tilde{\beta}_2$,
  respectively, from a multivariate normal distribution with mean
  equal to the previously estimated coefficient vector and covariance
  equal to the previously estimated covariance matrix.
\item Given values of $\tilde{\beta}_0$, $\tilde{\beta}_1$, and
  $\tilde{\beta}_2$, for each runner draw an error $\tilde{e}$ from a
  normal distribution with mean zero and a standard deviation equal to
  the standard deviation of the original regression model's residuals.
\item For each runner, predict her final result by combining the
  randomly generated beta coefficients and error terms,
  $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +
  \tilde{\beta}_2\,X_i^2 + \tilde{e}$.
\item Eliminate each runner from the simulated race with a probability
  equal to the observed fraction of marathoners who did not finish the
  marathon.
% \item Calculate the difference in time and difference in ranking
%   between Anna Hahner and Lisa Hahner, assuming both finished the
%   race.
\item Repeat above steps 10,000 times.
\end{enumerate}

From our simulations we can generate intervals which describe the
extent of the variability in marathon finishing times. For example, in
95\% of the simulations in which Anna Hahner completed the marathon,
her finishing time was between 9,480 seconds and 10,703 seconds. This
is consistent with Anna's observed finishing time in Rio, which was
9,932 seconds. In fact, our simulations estimate that Anna should have
finished the Rio marathon in 10,092 seconds, which is slightly more
than two and one-half minutes slower than her actual time. Lisa
Hahner's corresponding 95\% interval is 9,427 to 10,619 with an
observed finishing time of 9,933 seconds.

The bottom line here is that our simulated marathon finishes are
consistent with the Hahner twins' finishing times. And note that the
regression model underlying the simulations was estimated without the
Hahner twins (and same for the Kim twins and Luik triplets). According
to our simulations, then, neither Hahner twin ran appreciably slowly,
conditional on personal best times prior to the Rio Olympics.

@@

The histograms in Figure \ref{fig:simdiff} raise questions about the
credibility of Anna and Lisa Hahner's story. In particular, the
histogram suggest that a simultaneous finish in the Rio marathon would
be very rare if Anna and Lisa had run independently of each other. For
example, in fewer than 300 of the 10,000 simulated races did Anna and
Lisa finish within 30 seconds of each other. Moreover, they finished
in consecutive rank in fewer than 175 of the 10,000 simulated races.
The close finish that was observed, where Anna and Lisa crossed the
finish line one after the other, would have been highly unlikely if
the two German twins had raced independently of each other.

Parallel with our prior analyses, we repeated our simulations but this
time used half marathon splits as the predicting variable in our
simulations. The results of this are in Figure \ref{fig:simdiffhalf}.
While this use of marathon splits reduces the variation in the
predicted outcome of each runner and therefore reduces the expected
distance between Anna and Lisa Hahner, even with half marathon split
as a measure of ability it is still quite rare for Anna and Lisa to
finish simultaneously or consecutively.

@@

An important point regarding our simulations is the way that we
handled DNFs. As our description above indicates, we assumed that DNF
probabilities are the same for all runners and that the likelihood of
a DNF is a not function of a runner's anticipated marathon finishing
time. The rug marks in Figure \ref{fig:45degreeplot} suggest that
runners with better personal best times may be more likely to DNF than
other runners, \emph{ceteris paribus}. We suspect that this occurs
because some better runners may expend excessive energy trying to
achieve a good result in the marathon and in so doing injure or
exhaust themselves; lesser runners, in contrast, may be content to
finish respectably. Regardless of the validity of this conjecture,
Figure \ref{fig:45degreeplot} shows that the Hahner twins are
representative of the sort of runners who DNFed in Rio. Since our
simulated Hahner statistics are conditioned on both Hahner twins
finishing, it follows that they are conservative. The fact that both
women finished the marathon was notable in and of itself, and by
discounting the possibility of a Hahner DNF we are giving the benefit
of the doubt to the Hahner twins.
